Miniproject - Algebraic Geometry

### 2021-12-19 Algebraic geometry/submission.tex
small fixed, mainly to ANT
Final corrections
Finished alg geo

\title{Miniproject - Algebraic Geometry}
\author{Simon Pohmann}
\date{}

@@

For this document, let $V$ be a finitely dimensional vector space with basis $B = (b_1, ..., b_n)$.

@@

    Note that by definition, the vectors $b_{i_1} \wedge ... \wedge b_{i_d}$ for any $1 \leq i_1, ..., i_d \leq n$ span $\extpow^d V$.
    By Lemma~\ref{prop:basic_properties_exterior_product}, we can rewrite any of them (that is nonzero) to be of the above form, so clearly the $b_{i_1} \wedge ... \wedge b_{i_d}$ for $i_1 < ... < i_d$ span $\extpow^d V$.

@@

    For the embedding $\phi: \Gr(2, V) \to \extpow^2 V$ we have

@@

    Here $\mathrm{lt}(I)$ is the leading term ideal of $I$, i.e. the ideal generated by the leading terms of all $f \in I$, w.r.t. $\preceq$.
\end{proposition}
\begin{proof}
    See \cite{kreuzer}.
\end{proof}
To apply this, first of all we have to collect information about the leading term ideal of $\I(\Gr(2, n))$.
This is done in the following lemma.
\begin{lemma}
    \label{prop:leading_term_ideal_grassmanian}
    Define the graded reverse monomial ordering $\preceq$ on $R := k[x_{i, j} \ | \ i < j]$ where the variables $x_{i, j}$ are ordered lexicographically w.r.t. $(i, j)$, i.e.
    \begin{equation*}
        x_{i, j} \leq x_{u, v} \ :\Leftrightarrow \ (i, j) \leq_{\text{lex}} (u, v)
    \end{equation*}
    Moreover, let
    \begin{equation*}
        I := \langle x_{i, j}x_{u, v} + x_{i, v}x_{j, u} - x_{i, u}x_{j, v} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
    be the ideal defining $\phi(\Gr(d, V))$ that was considered above. Then
    \begin{equation*}
        \mathrm{lt}(I) = J := \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
\end{lemma}
\begin{proof}
    We show that the set $G = \{ x_{iv} x_{ju} - x_{iu} x_{jv} + x_{ij} x_{uv} \ | \ i < j < u < v \}$ is a Gröbner basis of $I$.
    The claim then follows by the properties of Gröbner basis.

@@

    However, observe that $S$ only depends on the relative order of $i, j, u, v, a, b, c, d$.
    All possible orders of those are already present if we consider
    \begin{equation*}
        1 \leq i < j < u < v \leq 8 \quad \text{and} \quad 1 \leq a < b < c < d \leq 8
    \end{equation*}
    In other words, it suffices to show that $G$ is a Gröbner basis in the case $n = 8$.
    By hand, this requires exactly the same computation as started above, but it can be very easily checked using Computer Algebra.
    I have done this, and the result is as expected.
\end{proof}
At first glance, this lemma above seems quite deep, but it really is a repetitive application of the theory of Gröbner basis.

Now we have to study the structure of monomials generating $J = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle$.
At first glance, this seemed to me to be much simpler than the above, but in fact it turns out not to be, as there is some subtle combinatorial structure involved.
Hence, we will first introduce the combinatorial framework in which we will work now.
\begin{definition}
    Consider the graph $G_n = (V_n, E_n)$ where $V_n = \{ (i, j) \ | \ 0 \leq i < j \leq n \}$ and
    \begin{equation*}
        E_n = \{ \{ (i, v), (j, u) \} \ | \ 0 \leq i < j < u < v \leq n \}
    \end{equation*}
    Let further
    \begin{equation*}
        s_n(d) := |\{ I \subseteq V_n \ | \ I \ \text{independent set of size $d$}\}|
    \end{equation*}
\end{definition}
It might be possible to see from the definition how independent sets in this graph are connected to monomials in $J$.
If not, the connection is explained in the proof of the next theorem.
\begin{proposition}
    Let $d$ be the largest integer such that $s_n(d) \neq 0$, i.e. the size of the largest independent set in $G_n$.
    Then
    \begin{equation*}
        \dim(\Gr(2, n)) = d - 1
    \end{equation*}
    and
    \begin{equation*}
        \deg(\Gr(2, n)) = s_n(d)
    \end{equation*}
\end{proposition}
\begin{proof}
    The idea is to apply Macaulay's basis theorem to get information about the Hilbert function of $\Gr(2, n)$.
    To this end, observe that all the monomials generating $\mathrm{lt}(\I(\Gr(2, n)))$ are square-free.
    Hence, we can partition all monomials not in $\mathrm{lt}(\I(\Gr(2, n)))$ into sets depending on which variables occur in them.
    Doing this shows that leading coefficient of the Hilbert polynomial depends only on the groups with the maximal count of monomials.
    These monomials then form a maximal independent set in $G_n$, which allows us to relate degree and dimension of $\Gr(2, n)$ to $s_n(d)$.

@@

    On the other hand, it is easy to observe that the 0-th layer
    \begin{equation*}
        I = \{ v \in V \ | \ l(v) = 0 \}
    \end{equation*}
    is indeed an independent set, as for all elements $(i, j) \in I$ have either $i = 1$ or $j = n$.
\end{proof}
\begin{corollary}
    Have $\dim(\Gr(2, n)) = 2n - 4$.
\end{corollary}
\begin{proof}
    Follows directly from the previous two statements.
\end{proof}
I also think that it is not too difficult to show that
\begin{equation*}
    s_n(2n - 3) = \frac {(2(n - 2))!} {(n - 2)! (n - 1)!}
\end{equation*}
which then shows the degree formula for the case $d = 2$.
However, this is not a lecture on graph theory or combinatorics, and I already spent much more time on this Miniproject than on the others.
So I will not do the proof here, but I think it is already very interesting to see those graphs and the connection of $\deg(\Gr(2, n))$ to combinatorial problems.
Finally, in the case of $n = 6$ we can simply count the independent sets by hand, and find the following result.
\begin{corollary}[Question (e)]
    Have $\dim(\Gr(2, 6)) = 8$ and $\deg(\Gr(2, 6)) = 14$.
\end{corollary}
\begin{proof}
    The dimension follows directly from the previous general case.
    For the degree, just count the maximal independent sets in $G_6$, as displayed in Example~\ref{ex:Gn}.
    This is slightly tricky, but it is not too hard to see that there are exactly 14 of them.
\end{proof}
Of course, the above statement on $\Gr(2, 6)$ can also easily be found using Computer Algebra.
For example, the following Sage script shows that
\begin{equation*}
    p_{\Gr(2, 6)} = \frac 1 {2880} t^8 + \frac 1 {120} t^7 + \frac {41} {480} t^6 + \frac {39} {80} t^5 + \frac {541} {320} t^4 + \frac {291} {80} t^3 + \frac {3401} {720} t^2 + \frac {101} {30} t + 1
\end{equation*}
from which we easily deduce that
\begin{equation*}
    \dim(\Gr(2, 6)) = 8 \quad \text{and} \quad \deg(\Gr(2, 6)) = \frac 1 {2880} \cdot 8! = 14
\end{equation*}
\lstinputlisting[language = python]{./gr26.sage}
\printbibliography
\end{document}

>>>>

\title{Miniproject - Algebraic Geometry}
\author{Student Number: 1462479}
\date{19.02.2021}

@@

\subsection*{Introduction}

In this Miniproject, we want to study the Grassmanian
\begin{equation*}
    \Gr(d, V) := \{ U \subseteq V \ | \ U \ \text{$d$-dimensional subspace} \}
\end{equation*}
as an object of Algebraic Geometry.
First of all, this requires finding a ``natural'' embedding of $\Gr(d, V)$ in to projective space, and showing that w.r.t. that embedding, the Grassmanian is a projective variety.

After that is established, we can continue to study its geometric properties, like vanishing ideal, dimension and degree.
This part will only be considered in the case $d = 2$.

\subsection*{Conventions}
We use the convention that $\N = \{ n \in \Z \ | \ n \geq 0 \}$.
Further, we write $a \divides b$ if $a$ divides $b$ and $a \perp b$ if $a$ and $b$ are coprime.
During this Miniproject, let $V$ be a finitely dimensional vector space with basis $B = (b_1, ..., b_n)$.

@@

    Note that by definition, the vectors $b_{i_1} \wedge ... \wedge b_{i_d}$ for $1 \leq i_1, ..., i_d \leq n$ span $\extpow^d V$.
    By Lemma~\ref{prop:basic_properties_exterior_product}, we can rewrite any of them (that is nonzero) to be of the above form, so clearly the $b_{i_1} \wedge ... \wedge b_{i_d}$ for $i_1 < ... < i_d$ span $\extpow^d V$.

@@

    For the embedding $\phi: \Gr(2, V) \to \Proj(\extpow^2 V)$ we have

@@

    Here $\mathrm{lt}(I)$ is the leading term ideal of $I$, i.e. the ideal generated by the leading terms of all $f \in I$, w.r.t. $\preceq$.
\end{proposition}
\begin{proof}
    See \cite{kreuzer}.
\end{proof}
To apply this, first of all we have to collect information about the leading term ideal of $\I(\Gr(2, n))$.
This is done in the following lemma.
\begin{lemma}
    \label{prop:leading_term_ideal_grassmanian}
    Define the graded reverse monomial ordering $\preceq$ on $R := k[x_{i, j} \ | \ i < j]$ where the variables $x_{i, j}$ are ordered lexicographically w.r.t. $(i, j)$, i.e.
    \begin{equation*}
        x_{i, j} \leq x_{u, v} \ :\Leftrightarrow \ (i, j) \leq_{\text{lex}} (u, v)
    \end{equation*}
    Moreover, let
    \begin{equation*}
        I := \langle x_{i, j}x_{u, v} + x_{i, v}x_{j, u} - x_{i, u}x_{j, v} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
    be the ideal defining $\phi(\Gr(d, V))$ that was considered above. Then
    \begin{equation*}
        \mathrm{lt}(I) = J := \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
\end{lemma}
\begin{proof}
    We show that the set $G = \{ x_{iv} x_{ju} - x_{iu} x_{jv} + x_{ij} x_{uv} \ | \ i < j < u < v \}$ is a Gröbner basis of $I$
    \footnote{In this proof, we write $x_{ij} := x_{i, j}$ to somewhat simplify notation.}.
    The claim then follows by the properties of Gröbner basis.

@@

    However, observe that $S$ only depends on the relative order of $i, j, u, v, a, b, c, d$.
    All possible orders of those are already present if we consider
    \begin{equation*}
        1 \leq i < j < u < v \leq 8 \quad \text{and} \quad 1 \leq a < b < c < d \leq 8
    \end{equation*}
    In other words, it suffices to show that $G$ is a Gröbner basis in the case $n = 8$.
    By hand, this requires exactly the same computation as started above, but it can be very easily checked using Computer Algebra.
    I have done this, and the result is as expected.
\end{proof}
At first glance, this lemma above seems quite deep, but it really is a repetitive application of the theory of Gröbner basis.
\begin{corollary}
    Using the notation of Prop.~\ref{prop:equations_grassmanian_2}, we get that
    \begin{equation*}
        \I(\Gr(2, V)) = I
    \end{equation*}
\end{corollary}
\begin{proof}
    By Hilbert's Nullstellensatz, it is sufficient to show that $I$ is radical, as we know from Prop.~\ref{prop:equations_grassmanian_2} that $\V(I) = \Gr(2, V)$.
    This follows from a very standard argument in Commutative Algebra, but for completeness, we will present the full proof here.

    Assume that $f^d \in I$ for some $f \in R$.
    We show that $f \in I$ by using induction on the position of $\mathrm{lt}(f)$ w.r.t. $\preceq$ (note that this is a well-ordering, so we can use induction on it).
    The base case $f \in k$ is trivial, so assume that $f$ is not constant.

    By Lemma~\ref{prop:leading_term_ideal_grassmanian} we see that
    \begin{equation*}
        \mathrm{lt}(f)^d = \mathrm{lt}(f^d) \in J = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle
    \end{equation*}
    Hence
    \begin{equation*}
        \mathrm{lt}(f)^d \equiv x_{i, v} x_{j, u}
    \end{equation*}
    for some $i < j < u < v$.
    Clearly $x_{i, v} \perp x_{j, u}$, and it follows that
    \begin{equation*}
        x_{i, v} x_{j, u} \divides \mathrm{lt}(f)
    \end{equation*}
    Now consider
    \begin{equation*}
        f' := f - \frac {\mathrm{lt}(f)} {x_{i, v} x_{j, u}} (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v}) \in R
    \end{equation*}
    Clearly $\mathrm{lt}(f') \prec \mathrm{lt}(f)$ as $x_{i, v} x_{j, v}, x_{i, j} x_{u, v} \prec x_{i, v} x_{j, u} \equiv \mathrm{lt}(f)$.
    Furthermore
    \begin{equation*}
        (f')^d = \sum_{l = 0}^d {d \choose i} f^i (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v})^{d - i} \in I
    \end{equation*}
    so by induction hypothesis, find that $f' \in I$.
    It follows that also
    \begin{equation*}
        f = f'  - \frac {\mathrm{lt}(f)} {x_{i, v} x_{j, u}} (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v}) \in I
    \end{equation*}
    and the claim is shown.
\end{proof}
Now we have to study the structure of monomials generating $J = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle$.
At first glance, this seemed to me to be much simpler than the above, but in fact it turns out not to be, as there is some subtle combinatorial structure involved.
Hence, we will first introduce the combinatorial framework in which we will work now.
\begin{definition}
    Consider the graph $G_n = (V_n, E_n)$ where $V_n = \{ (i, j) \ | \ 0 \leq i < j \leq n \}$ and
    \begin{equation*}
        E_n = \{ \{ (i, v), (j, u) \} \ | \ 0 \leq i < j < u < v \leq n \}
    \end{equation*}
    Let further
    \begin{equation*}
        s_n(d) := |\{ I \subseteq V_n \ | \ I \ \text{independent set of size $d$}\}|
    \end{equation*}
\end{definition}
It might be possible to see from the definition how independent sets in this graph are connected to monomials in $J$.
If not, the connection is explained in the proof of the next theorem.
\begin{proposition}
    Let $d$ be the largest integer such that $s_n(d) \neq 0$, i.e. the size of the largest independent set in $G_n$.
    Then
    \begin{equation*}
        \dim(\Gr(2, n)) = d - 1
    \end{equation*}
    and
    \begin{equation*}
        \deg(\Gr(2, n)) = s_n(d)
    \end{equation*}
\end{proposition}
\begin{proof}
    The idea is to apply Macaulay's basis theorem to get information about the Hilbert function of $\Gr(2, n)$.
    To this end, observe that all the monomials generating $\mathrm{lt}(\I(\Gr(2, n)))$ are square-free.
    Hence, we can partition all monomials not in $\mathrm{lt}(\I(\Gr(2, n)))$ into sets depending on which variables occur in them.
    Doing this shows that leading coefficient of the Hilbert polynomial depends only on the groups with the maximal count of monomials.
    These monomials then form a maximal independent set in $G_n$, which allows us to relate degree and dimension of $\Gr(2, n)$ to $s_n(d)$.

@@

    On the other hand, it is easy to observe that the 0-th layer
    \begin{equation*}
        I = \{ v \in V \ | \ l(v) = 0 \}
    \end{equation*}
    is indeed an independent set, as for all elements $(i, j) \in I$ have either $i = 1$ or $j = n$.
\end{proof}
\begin{corollary}
    Have $\dim(\Gr(2, n)) = 2n - 4$.
\end{corollary}
\begin{proof}
    Follows directly from the previous two statements.
\end{proof}
I also think that it is not too difficult to show that
\begin{equation*}
    s_n(2n - 3) = \frac {(2(n - 2))!} {(n - 2)! (n - 1)!}
\end{equation*}
which then shows the degree formula for the case $d = 2$.
However, this is not a lecture on graph theory or combinatorics, and I already spent much more time on this Miniproject than on the others.
So I will not do the proof here, but I think it is already very interesting to see those graphs and the connection of $\deg(\Gr(2, n))$ to combinatorial problems.
Finally, in the case of $n = 6$ we can simply count the independent sets by hand, and find the following result.
\begin{corollary}[Question (e)]
    Have $\dim(\Gr(2, 6)) = 8$ and $\deg(\Gr(2, 6)) = 14$.
\end{corollary}
\begin{proof}
    The dimension follows directly from the previous general case.
    For the degree, just count the maximal independent sets in $G_6$, as displayed in Example~\ref{ex:Gn}.
    This is slightly tricky, but it is not too hard to see that there are exactly 14 of them.
\end{proof}
Of course, the above statement on $\Gr(2, 6)$ can also easily be found using Computer Algebra.
For example, the following Sage script shows that
\begin{equation*}
    p_{\Gr(2, 6)} = \frac 1 {2880} t^8 + \frac 1 {120} t^7 + \frac {41} {480} t^6 + \frac {39} {80} t^5 + \frac {541} {320} t^4 + \frac {291} {80} t^3 + \frac {3401} {720} t^2 + \frac {101} {30} t + 1
\end{equation*}
from which we easily deduce that
\begin{equation*}
    \dim(\Gr(2, 6)) = 8 \quad \text{and} \quad \deg(\Gr(2, 6)) = \frac 1 {2880} \cdot 8! = 14
\end{equation*}
\lstinputlisting[language = python]{./gr26.sage}

\subsection*{Conclusion}
We have seen that the linear dependency structure between $d$-element subsets of $V$ can be described very well using the external power $\extpow^d V$.
Therefore, this might look like a good candidate for describing the Grassmanian, and in this Miniproject, we have seen that indeed it is.
In particular, it can provide a natural projective space in which to embed the Grassmanian, so that it becomes a projective variety.

After this embedding is constructed, we have shown its close connection to the determinant, which is also a good way to describe linear dependency structures.
Notably, the determinant is linear in each column or row, and the external product $\cdot \wedge ... \wedge \cdot$ is linear in each component.
After that, we restricted to the case $d = 2$ and looked at the properties of $\Gr(2, d)$ as projective variety.
Here we could show that the variety is cut out by quadratic equations, and analyse the concrete varieties $\Gr(2, 4), \Gr(2, 5)$ and $\Gr(2, 6)$ in detail.

Finally, we have used some techniques from the theory of Gröbner basis in Commutative Algebra to further study the general case of $\Gr(2, n)$.
These methods are quite useful to handle many multivariate polynomials, and in particular the cancellations that can occur when considering $R$-combinations of them.
As a result of this part, we could determine the vanishing ideal of $\Gr(2, n)$.
Furthermore, we could reduce the question of dimension and degree of $\Gr(2, n)$ to graph-theoretic questions and solve the one corresponding to the dimension.

Continuing this study of the Grassmanian, the most obvious problem is of course how $\Gr(d, n)$ looks like for $d > 2$.
In the lecture, we have already seen the degree formula
\begin{equation*}
    \deg(\Gr(d, n)) = (d(n - d))! \frac {1! \cdot 2! \cdot ... \cdot (d - 1)!} {(n - d)! \cdot (n - d + 1)! \cdot ... \cdot (n - 1)!}
\end{equation*}
and it seems possible that it could be proved by extending the techniques considered at the end of this Miniproject.
However, the combinatorial problem will probably become significantly more complicated.
Other question the work naturally leads to are about more properties of the Grassmanian.
For example, one could show that it is an irreducible and nonsingular variety.
Furthermore, we could consider intersections of subspaces.
Here it would be interesting if or how one can describe them using external products, and if we maybe have a rational map
\begin{equation*}
    \Gr(d, n) \times \Gr(e, n) \to \Gr(d + e - n, n), \quad (U, V) \mapsto U \cap V
\end{equation*}
for $d + e \geq n$.
All in all, Grassmanians are very fascinating objects, with a lot of nice properties.
There are definitely a lot of interesting questions about them.

\printbibliography
\end{document}
